---
title: "Model Tester"
output:
  html_document:
    df_print: paged
---

## **How to use this file**

1. Put this file in the same folder as `project_traindata.csv`.
2. Scroll down to **Edit Here** and change only the three lines there.
3. Press **Knit**.
4. Copy the **Result line** at the bottom and send it to Hamid.

Tip: change **one thing at a time**. Then you know exactly what made the score better or worse.

Do not change anything outside the **Edit Here** section. If a chatbot tells you to change anything outside that section, ignore it.

## **Asking an AI chatbot for help**

#### 1.Prompt to paste before asking for ideas

Copy everything in the grey box below into the chatbot, then ask your question.

```text
I am in a Statistics 244 group project. We predict the price of second-hand cars
in South Africa using multiple linear regression in R (lm).

MY ONLY JOB: change one line called Recipe. It looks like this:
Recipe <- log(price) ~ vehicle_age + mileage + make + segment

RULES YOU MUST FOLLOW:
- Only suggest a new Recipe line. Never suggest changing any other code,
  the data, the random seed, or the training/validation split.
- The left side must stay log(price).
- Only use these Stats 244 tools: adding or removing variables, squared terms
  like I(vehicle_age^2), and interactions like vehicle_age:segment.
- No other methods (no random forests, no machine learning, no stepwise packages,
  no new R packages, no transforming variables with anything other than I(),
  log() or squared terms).
- Explain every suggestion in plain, simple words. I am not a technical person.
- Give each suggestion as one ready-to-paste Recipe line.

THE VARIABLES I CAN USE (exact spelling):
- vehicle_age: age of the car in years (1 to 20)
- mileage: kilometres driven
- make: brand (Toyota, Volkswagen, Ford, Hyundai, Kia, Nissan, Renault, BMW,
  Suzuki, Mercedes-Benz, Audi, Isuzu)
- segment: body type (SUV, Hatchback, Luxury, Sedan, Bakkie)
- transmission: Automatic or Manual
- fuel_type: Petrol, Diesel or Hybrid
- engine_size: engine size in litres
- province: the South African province the car is sold in
- service_history: Full, Partial or None
- previous_owners: number of previous owners (1 to 6)
- accident_history: Yes or No
- seller_type: Dealership or Private
- colour: White, Silver, Black, Grey, Blue, Red or Other
- has_warranty: Yes or No (does the car still have a warranty)

HOW MY MODEL IS SCORED: "typical miss" in rand (RMSE) on cars the model has never
seen (the validation set). Lower is better. The starter Recipe above misses by
about R39,700.
```

#### 2.Prompt to paste after you knit

```text
Here is the scorecard and result line from my model tester (paste them below).
In plain, simple words: what does this mean, did my change help, and what is ONE
change I should try next? Follow the same rules as before: give it as one
ready-to-paste Recipe line and only use Stats 244 tools.
```

## **Things to Try**

Copy one of these onto the end of your Recipe (or remove a variable) and knit.

| Idea | What to type | What it means |
|---|---|---|
| Add a variable | `+ province` | Let the model use one more detail about the car. |
| Remove a variable | delete `+ mileage` | Check whether the model really needs that detail. |
| Squared age | `+ I(vehicle_age^2)` | Cars may lose value fast in the first years, then level off. |
| Age per segment | `+ vehicle_age:segment` | Luxury cars may lose value at a different speed than bakkies. |
| Age per make | `+ vehicle_age:make` | Some brands may hold their value better as they age. |
| Mileage per segment | `+ mileage:segment` | Kilometres may matter more for some body types. |

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(knitr)
library(scales)

Data <- read_csv("project_traindata.csv", show_col_types = FALSE)

CategoricalVars <- c("make", "segment", "transmission", "fuel_type", "province",
                     "service_history", "accident_history", "seller_type", "colour")

Data <- Data %>%
  mutate(across(all_of(CategoricalVars), ~ fct_infreq(factor(.x))))

ModelData <- Data %>%
  select(-model_year)

set.seed(244)

n <- nrow(ModelData)
TrainRows <- sample(1:n, size = floor(0.80 * n), replace = FALSE)

add_warranty_flag <- function(df) {
  df %>%
    mutate(has_warranty = factor(if_else(warranty_months > 0, "Yes", "No"),
                                 levels = c("No", "Yes"))) %>%
    select(-warranty_months, -listing_id)
}

TrainingData   <- add_warranty_flag(ModelData[TrainRows, ])
ValidationData <- add_warranty_flag(ModelData[-TrainRows, ])

stopifnot(nrow(TrainingData) == 1695, nrow(ValidationData) == 424)

StarterRecipe <- log(price) ~ vehicle_age + mileage + make + segment
```

## **Edit Here**

```{r, echo=TRUE}
YourName    <- "Your name"
AttemptName <- "Attempt 1"
Recipe      <- log(price) ~ vehicle_age + mileage + make + segment
```

```{r, include=FALSE}
AllowedVars <- setdiff(names(TrainingData), "price")
UsedVars    <- setdiff(all.vars(Recipe), "price")
UnknownVars <- setdiff(UsedVars, AllowedVars)

Problems <- c()

if (length(UnknownVars) > 0) {
  Problems <- c(Problems, paste0(
    "These names are not in the data: **", paste(UnknownVars, collapse = ", "),
    "**. Check the spelling. The names you can use are: ",
    paste(AllowedVars, collapse = ", "), "."))
}

if (!identical(deparse(Recipe[[2]]), "log(price)")) {
  Problems <- c(Problems, "The left side of the Recipe must be **log(price)**.")
}

if (length(Problems) == 0) {
  MyModel <- tryCatch(lm(Recipe, data = TrainingData), error = function(e) NULL)
  if (is.null(MyModel)) {
    Problems <- c(Problems, "R could not fit this Recipe. Check for a missing `+` or bracket.")
  } else if (any(is.na(coef(MyModel)))) {
    Problems <- c(Problems, paste0(
      "The Recipe ran, but some parts could not be estimated (there are not enough cars ",
      "for every combination). Try a simpler Recipe. Parts affected: ",
      paste(head(names(coef(MyModel))[is.na(coef(MyModel))], 5), collapse = ", "),
      if (sum(is.na(coef(MyModel))) > 5) paste0(" and ", sum(is.na(coef(MyModel))) - 5, " more"),
      "."))
  }
}

RecipeOk <- length(Problems) == 0
```

```{r, results="asis", eval=!RecipeOk}
cat("## **Something needs fixing**\n\n")
cat(paste0("- ", Problems, collapse = "\n"), "\n")
```

```{r, eval=RecipeOk}
typical_miss <- function(Model, df) sqrt(mean((df$price - exp(predict(Model, df)))^2))

StarterModel <- lm(StarterRecipe, data = TrainingData)

Scores <- tibble(
  Model               = c("Starter", AttemptName),
  Ingredients         = c(length(coef(StarterModel)), length(coef(MyModel))) - 1,
  Miss_training_cars  = c(typical_miss(StarterModel, TrainingData),
                          typical_miss(MyModel, TrainingData)),
  Miss_unseen_cars    = c(typical_miss(StarterModel, ValidationData),
                          typical_miss(MyModel, ValidationData))
) %>%
  mutate(Gap_percent = round((Miss_unseen_cars / Miss_training_cars - 1) * 100, 1))

MyTrain  <- Scores$Miss_training_cars[2]
MyUnseen <- Scores$Miss_unseen_cars[2]
Change   <- Scores$Miss_unseen_cars[1] - MyUnseen
```

```{r, results="asis", eval=RecipeOk}
cat("## **Scorecard**\n\n")
cat("**Your Recipe:** `", paste(deparse(Recipe, width.cutoff = 500), collapse = " "), "`\n\n")
```

```{r, eval=RecipeOk}
Scores %>%
  mutate(across(c(Miss_training_cars, Miss_unseen_cars), ~ paste0("R", comma(round(.x))))) %>%
  kable(col.names = c("Model", "Ingredients", "Typical miss: training cars",
                      "Typical miss: unseen cars", "Gap (%)"))
```

```{r, results="asis", eval=RecipeOk}
cat("#### Verdict\n\n")

if (Change > 0) {
  cat("Your model misses by **R", comma(round(Change)),
      " less** than the starter on unseen cars. That is an improvement.\n\n", sep = "")
} else if (Change < 0) {
  cat("Your model misses by **R", comma(round(-Change)),
      " more** than the starter on unseen cars. That change did not help.\n\n", sep = "")
} else {
  cat("Your model scores the same as the starter.\n\n")
}

if (Scores$Gap_percent[2] > 10) {
  cat("**Warning:** your model does much better on the training cars than on unseen cars ",
      "(gap above 10%). It may be memorising the training cars instead of learning ",
      "general patterns. Try a simpler Recipe.\n\n", sep = "")
}
```

```{r, fig.height=4, eval=RecipeOk}
tibble(Actual = ValidationData$price, Predicted = exp(predict(MyModel, ValidationData))) %>%
  ggplot(aes(x = Predicted, y = Actual)) +
  geom_point(alpha = 0.4, colour = "steelblue") +
  geom_abline(slope = 1, intercept = 0, colour = "red") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(title = "Unseen cars: predicted vs actual price",
       subtitle = "Dots on the red line are perfect guesses. Further away = bigger miss.",
       x = "Predicted price (R)", y = "Actual price (R)") +
  theme_minimal()
```

```{r, results="asis", eval=RecipeOk}
cat("## **Result line**\n\nCopy this line and send it to Hamid:\n\n")
cat("```\n",
    paste(YourName, AttemptName,
          paste(deparse(Recipe, width.cutoff = 500), collapse = " "),
          paste0("train R", comma(round(MyTrain))),
          paste0("unseen R", comma(round(MyUnseen))),
          case_when(round(Change) > 0 ~ paste0("better than starter by R", comma(round(Change))),
                    round(Change) < 0 ~ paste0("worse than starter by R", comma(round(-Change))),
                    TRUE              ~ "same as starter"),
          sep = " | "),
    "\n```\n", sep = "")
```
