---
title: "step_0-Abdul-Hamid Rabiu"
output:
  html_document:
    df_print: paged
---

## **Global**

This chunk loads the packages, reads the training data and stores the names of the
numerical and categorical variables in two vectors. Every function further down works
off these vectors, so a variable only has to be listed once. `listing_id` is left out
of both vectors because it identifies an advertisement and is not a vehicle
characteristic.

```{r, include=FALSE}
set.seed(123)

library(readr)
library(dplyr)
library(tidyverse)
library(knitr)
library(ggplot2)
library(scales)

Data <- read_csv("project_traindata.csv")

NumericVars <- c("price", "model_year", "vehicle_age", "mileage",
                 "engine_size", "previous_owners", "warranty_months")

CategoricalVars <- c("make", "segment", "transmission", "fuel_type", "province",
                     "service_history", "accident_history", "seller_type", "colour")
```

## **Data Checks**

Before looking at any single variable it is worth checking the shape of the data set,
whether any values are missing and whether each advertisement appears only once.
Missing values matter because a regression drops any row with a missing value in a
variable that is in the model, so a variable with many missing entries can quietly
shrink the sample.

```{r}
dim(Data)

kable(
  data.frame(Missing = sapply(Data, function(x) sum(is.na(x)))),
  caption = "Missing values per variable"
)

n_distinct(Data$listing_id) == nrow(Data)
```

## **Helper Functions**

Rather than repeating the same code sixteen times, the functions below take a variable
name as text and return one piece of output each. `num_table` gives the descriptive
statistics together with a 95% confidence interval for the mean, `num_hist` shows the
shape of the distribution, `num_scatter` shows the relationship with `price`, and
`num_scatter_segment` splits that same relationship by vehicle segment so that a
difference in slope between segments becomes visible.

`num_table` is written with base R functions only (`mean`, `sd`, `median`, `quantile`,
`min`, `max`) so that skewness, spread and the confidence interval for the mean are all
in one table without depending on an extra package.

```{r, include=FALSE}
skewness <- function(x) {
  n <- length(x)
  m <- mean(x)
  s <- sd(x)
  (sum((x - m)^3) / n) / s^3
}

num_table <- function(var) {
  x <- Data %>% pull(.data[[var]])
  ci <- t.test(x)$conf.int

  stats <- data.frame(
    n = length(x),
    mean = mean(x),
    sd = sd(x),
    median = median(x),
    min = min(x),
    max = max(x),
    Q1 = quantile(x, 0.25),
    Q3 = quantile(x, 0.75),
    skew = skewness(x),
    CI_lower = ci[1],
    CI_upper = ci[2]
  )
  rownames(stats) <- var
  kable(round(stats, 2), caption = paste("Descriptive statistics:", var))
}

num_hist <- function(var, bins = 30) {
  ggplot(Data, aes(x = .data[[var]])) +
    geom_histogram(bins = bins, fill = "steelblue", colour = "white") +
    scale_x_continuous(labels = comma) +
    labs(title = paste("Distribution of", var), x = var, y = "Count") +
    theme_minimal()
}

num_scatter <- function(var) {
  ggplot(Data, aes(x = .data[[var]], y = price)) +
    geom_point(alpha = 0.3) +
    geom_smooth(method = "loess", se = FALSE, colour = "red") +
    scale_x_continuous(labels = comma) +
    scale_y_continuous(labels = comma) +
    labs(title = paste("price vs", var), x = var, y = "price") +
    theme_minimal()
}

num_scatter_segment <- function(var) {
  ggplot(Data, aes(x = .data[[var]], y = price, colour = segment)) +
    geom_point(alpha = 0.3) +
    geom_smooth(method = "lm", se = FALSE) +
    scale_x_continuous(labels = comma) +
    scale_y_continuous(labels = comma) +
    labs(title = paste("price vs", var, "by segment"), x = var, y = "price") +
    theme_minimal()
}
```

The categorical functions follow the same idea. `cat_table` gives absolute and relative
frequencies, `cat_price_table` gives the price summary within each category,
`cat_bar` shows how many observations fall in each category, and `cat_box` shows how
the price distribution differs between categories.

```{r, include=FALSE}
cat_table <- function(var) {
  Data %>%
    count(.data[[var]]) %>%
    mutate(Percent = round(n / sum(n) * 100, 1)) %>%
    rename(Category = 1, Frequency = n) %>%
    kable(caption = paste("Frequencies:", var))
}

cat_price_table <- function(var) {
  Data %>%
    group_by(.data[[var]]) %>%
    summarise(
      number_of_vehicles = n(),
      Average_price = round(mean(price), 0),
      Median_price = round(median(price), 0),
      spread_in_group = round(sd(price), 0),
      Min_price = min(price),
      Max_price = max(price)
    ) %>%
    rename(Category = 1) %>%
    kable(caption = paste("price by", var))
}

cat_bar <- function(var) {
  ggplot(Data, aes(x = fct_infreq(as.factor(.data[[var]])))) +
    geom_bar(fill = "steelblue") +
    labs(title = paste("Number of vehicles per", var), x = var, y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

cat_box <- function(var) {
  ggplot(Data, aes(x = as.factor(.data[[var]]), y = price)) +
    geom_boxplot(fill = "steelblue", alpha = 0.6) +
    scale_y_continuous(labels = comma) +
    labs(title = paste("price by", var), x = var, y = "price") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## **Numerical Variables**

Each variable below gets the same three outputs: a table of descriptive statistics, a
histogram and a scatterplot against `price`. Read them together and ask whether the
distribution is symmetric or skewed, whether there are impossible or extreme values,
whether the relationship with price rises or falls, and whether the loess curve is close
to a straight line or clearly bends. A bend is the signal that a squared term or a
transformation may be worth trying later.

#### 1.price

`price` is the response, so its shape decides whether the model should be fitted on
`price` or on a transformation such as `log(price)`. The second plot repeats the
histogram on the log scale for comparison.

```{r}
num_table("price")
num_hist("price")

ggplot(Data, aes(x = log(price))) +
  geom_histogram(bins = 30, fill = "darkorange", colour = "white") +
  labs(title = "Distribution of log(price)", x = "log(price)", y = "Count") +
  theme_minimal()
```

#### 2.model_year

```{r}
num_table("model_year")
num_hist("model_year", bins = 20)
num_scatter("model_year")
```

#### 3.vehicle_age

```{r}
num_table("vehicle_age")
num_hist("vehicle_age", bins = 20)
num_scatter("vehicle_age")
num_scatter_segment("vehicle_age")
```

#### 4.mileage

```{r}
num_table("mileage")
num_hist("mileage")
num_scatter("mileage")
num_scatter_segment("mileage")
```

#### 5.engine_size

```{r}
num_table("engine_size")
num_hist("engine_size", bins = 26)
num_scatter("engine_size")
num_scatter_segment("engine_size")
```

#### 6.previous_owners

`previous_owners` is a count that only takes a few whole-number values, so a bar chart
and a boxplot of price per number of owners are easier to read than a histogram. This
also shows whether it behaves more naturally as a numeric variable or as a factor.

```{r}
num_table("previous_owners")
cat_bar("previous_owners")
cat_box("previous_owners")
```

#### 7.warranty_months

Count how many vehicles have no warranty left before reading the histogram, because a
large block of zeros changes how the rest of the distribution should be interpreted and
may suggest an indicator variable for "has warranty" alongside the number of months.

```{r}
num_table("warranty_months")
sum(Data$warranty_months == 0)
num_hist("warranty_months", bins = 30)
num_scatter("warranty_months")
```

## **Relationship between model_year and vehicle_age**

These two variables are built from the same information, so including both can create
multicollinearity. The correlation and the scatterplot show how closely they track each
other, and the last line checks whether the two always add up to the same reference year.

```{r}
cor(Data$model_year, Data$vehicle_age)

ggplot(Data, aes(x = model_year, y = vehicle_age)) +
  geom_point(alpha = 0.4) +
  labs(title = "vehicle_age vs model_year", x = "model_year", y = "vehicle_age") +
  theme_minimal()

table(Data$model_year + Data$vehicle_age)
```

## **Correlation Between Numerical Variables**

A correlation matrix gives a quick overview of which numerical predictors move with
`price` and which move with each other. Strong correlation between two predictors is a
warning for multicollinearity later on.

```{r}
Data %>%
  select(all_of(NumericVars)) %>%
  cor() %>%
  round(2) %>%
  kable(caption = "Correlation matrix of numerical variables")
```

## **Categorical Variables**

Each variable below gets a frequency table, a price summary per category, a bar chart
and a boxplot. Look at whether any category has very few observations, since small
categories give unstable estimates and are candidates for being grouped into an "Other"
level. Then compare the boxplots: if the price distributions sit at clearly different
levels, the variable is likely to help predict price.

#### 1.make

```{r}
cat_table("make")
cat_price_table("make")
cat_bar("make")
cat_box("make")
```

#### 2.segment

```{r}
cat_table("segment")
cat_price_table("segment")
cat_bar("segment")
cat_box("segment")
```

#### 3.transmission

```{r}
cat_table("transmission")
cat_price_table("transmission")
cat_box("transmission")
```

#### 4.fuel_type

```{r}
cat_table("fuel_type")
cat_price_table("fuel_type")
cat_bar("fuel_type")
cat_box("fuel_type")
```

#### 5.province

```{r}
cat_table("province")
cat_price_table("province")
cat_bar("province")
cat_box("province")
```

#### 6.service_history

```{r}
cat_table("service_history")
cat_price_table("service_history")
cat_bar("service_history")
cat_box("service_history")
```

#### 7.accident_history

```{r}
cat_table("accident_history")
cat_price_table("accident_history")
cat_box("accident_history")
```

#### 8.seller_type

```{r}
cat_table("seller_type")
cat_price_table("seller_type")
cat_box("seller_type")
```

#### 9.colour

```{r}
cat_table("colour")
cat_price_table("colour")
cat_bar("colour")
cat_box("colour")
```

## **Categorical Variables Within Segment**

The project notes that the effect of a predictor may differ between vehicle segments.
Splitting a boxplot by segment shows whether the gap between categories is roughly the
same in every segment or whether it changes, which is what an interaction term would
capture.

```{r}
cat_box("transmission") + facet_wrap(~ segment)
cat_box("fuel_type") + facet_wrap(~ segment)
```

## **Notes For Meeting 2**

Write the three to five sentence conclusion for each variable here, one short paragraph
per variable, each supported by the table or plot above. Then list three findings from
the data and one model idea, such as a transformation of the response, a squared term
for a predictor whose loess curve bends, or an interaction with `segment`.
